---
title: "Age and Time in Package 'dembase'"
author: "John Bryant"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Age and Time in Package 'dembase'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Labels

Everything denominated in years


## Labelling points

```{r, eval = TRUE, echo = FALSE}
date <- seq(from = as.Date("2000-01-01"), by = "1 month", length.out = 4)
frac_year <- round(as.integer(date) / as.integer(as.Date("2001-01-01") - as.Date("2000-01-01")), 3)
```

| Value         | Dimension | Length | Rule           | Example                  |
|:------------  |:--------- |:------ |:-------------- |:------------------------ |	
| Start day     | Time      | $\ge 2$| Year-Month-Day | `2000, 2000.003` $\to$ `"2000-01-01", "2000-01-02"` |
| Start month   | Time      | $\ge 2$| Year-Month     | `2000, 2000.085` $\to$ `"2000-01-01", "2000-02-01` |
| Start quarter | Time      | $\ge 2$| Year-Quarter   | `2000, 2000.249` $\to$ `"2000-01-01", "2000-04-01"` |
| Any           | Any       | Any    | Value          | `80.1` $\to$ `"80.1"`    |
----------------------------------------------------------------------------------



If every point is the start of a day, and more than one point, then label using date, 
If every point is the start of a month, and more than one point, then label using date. Eg `r 2000 + frac_year` is displayed as `r date`.

Otherwise, label with number.


## Labelling intervals


| Lower limit | Upper limit | Dimtype | Rule      | Example       |
|:------------|:------------|:----------|:----------|:--------------|	
| Integer | Lower+1 | age |  Lower | `5, 6` $\to$ `"5"` |
| Integer | Lower+1 | time, cohort | Lower or Upper | `5, 6` $\to$ `"5"` or `"6"` |
| Integer | Integer | age | Lower-\<Upper-1\> | `5, 10` $\to$ `"5-9"` |
| Start day | Start next day | time, cohort | Year-Month-Day | `2000, 2000.003` $\to$ `"2000-01-01"` |
| Start month | Start next month | time, cohort | Year-Month | `2000, 2000.085` $\to$ `"2000-01"` |
| Start quarter | Start next quarter | time, cohort | Year-Quarter | `2000, 2000.249` $\to$ `"2000-Q1"` |
| Any | Infinite | Any | Lower+ | `100, Inf` $\to$ `"100+"` |
| Infinite | Any | Any | <Upper | `-Inf, 2000` $\to$ `"<2000"` |
| Any | Any | Any | Lower-Upper | `2005, 2010` $\to$ `"2005-2010"` |
-------------------------------------------------------------------------

Only time and cohort can have labelStart = FALSE. Cannot have labelStart = FALSE if last interval is open.

Whether a one-year time interval is displayed using lower or upper depends on `labelStart` argument.

Only use "2000-01" labelling if every interval can be interpreted as a month and if months successively.  Similarly, only use "2000-Q1" labelling if every interval can be interpreted as a quarter.




# Cleaning data

cleanAgeGroup(age)
cleanPeriod(period)


## Creating labels from dates

dateToAgeGroup(date, dob, breaks, firstOpen = TRUE, lastOpen = TRUE)
dateToCohort(date, breaks, firstOpen = TRUE, lastOpen = FALSE)
dateToPeriod(date, breaks, firstOpen = FALSE, lastOpen = FALSE)
dateToTriangle(date, dob, breaks, firstOpen = TRUE, lastOpen = FALSE)


## Creating labels from years

ageToAgeGroup(age, breaks = seq(0, 100, 5), firstOpen = FALSE, lastOpen = TRUE) 
yearToPeriod(year, breaks, firstOpen = FALSE, lastOpen = FALSE)



## subarray

```
subarray(time > 2010)
subarray(time > "2010")
```

## Counts, accounts

Movements(population, births = NULL, internal = NULL, entries = list(), 
    exits = list(), net = list(), labelStart = TRUE)

Counts(object, dimtypes = NULL, dimscales = NULL, type = NULL)

exposure((object, triangles = FALSE, labelStart = TRUE)



